---
title: "10x_GBM example"
output:
  html_document:
    self_contained: true
---

# Prepare inputs
## Install packages (run once)
```{r Install packages (run once)}
# install.packages("BiocManager")
# BiocManager::install("EBImage")
# BiocManager::install("GenomicAlignments")
# BiocManager::install("clusterProfiler")
# devtools::install_github("cellgeni/visutils")
# devtools::install_github("mamarkevi/plotCoverage")
# devtools::install_github("cellgeni/scsajr")
```

## Load packages
```{r Load packages, message = FALSE}
library(Seurat)
library(plotCoverage)
library(clusterProfiler)
library(Matrix)
library(SummarizedExperiment)
library(scsajr)
```

## Set working directory
```{r Set working directory, cache = TRUE}
data_path <- "." # path to tutorial folder
setwd(data_path)
```

## Load reference
```{r Load gene description, cache = TRUE}
# gene_descr[dataframe]: gene descriptions
gene_descr_path <- "./nf-scsajr/ref/human_2020A/functional_annotation/gene.descr.rds"
gene_descr <- readRDS(gene_descr_path)
str(gene_descr)
# View(gene_descr)
```

```{r Load GTF (hidden), cache = TRUE}
# gtf[dataframe]: GTF file with gene annotations
gtf_path <- "./nf-scsajr/ref/human_2020A/gtf.rds"
gtf <- readRDS(gtf_path)
str(gtf)
# View(gtf)  # Too large to view
```


## Load input data
```{r Load input data, cache = TRUE}
# Sample -> BAM file (scRNA data mapped to reference genome)
# echo -e "sample\t$(pwd)/alignment.bam" > samples.tsv
samples <- read.table("./input/samples.tsv", col.names = c("sample_id", "bam_path"))

# Sample -> barcode, celltype
# awk -F, 'BEGIN { OFS="\t" }; NR>1 {print  "sample",$1,$2}' analysis/clustering/graphclust/clusters.csv > barcodes.tsv
barcodes <- read.table("./input/barcodes.tsv", col.names = c("sample_id", "barcode", "celltype"))
rownames(barcodes) <- paste0(barcodes$sample_id, "|", barcodes$barcode)

# Definition of splicing segments
segs_path <- "./nf-scsajr/ref/human_2020A/segments.csv"
segs <- read.csv(segs_path, row.names = 1)
```

---  
  
# Inspect pipeline output

## Load Pseudobulk Splicing Data
```{r Load Pseudobulk Splicing Data, cache = TRUE}
# pbas.rds: Pseudobulk AS segments data for all segments
# pb_as_filtered.rds [SummerizedExperiments]: Pseudobulk AS segments data filtered by coverage
pbasf_path <- "./nf-scsajr_output/rds/pb_as_filtered.rds"
pbasf <- readRDS(pbasf_path)

slotNames(pbasf)
```

| Column            | Meaning                                                                  |
| ----------------- | ------------------------------------------------------------------------ |
| `Rows`            | Splicing segments                                                        |
| `Cols`            | (sample, cell-type) pseudobulks                                          |
| `Assays`          | `i`: inclusion junction, `e`: exclusion junction, `psi`: percent spliced |
| `Metadata`        | Differential-splicing test results                                       |

<br><br>

### Row: Information about segments (features)
| Column            | Meaning                                                          |
| ----------------- | ---------------------------------------------------------------- |
| `feature_id`      | unique ID from Ensembl gene ID “ENSG00000188976.s7”              |
| `gene_id`         | Ensembl gene ID (e.g. “ENSG00000188976”)                         |
| `type`            | segment type (e.g. “EXN” for exon, “ALT” for alternative region) |
| `position`        | location within transcript (e.g. “INTERNAL”)                     |
| `sites`           | splice‐site pattern (e.g. “ad” = acceptor–donor)                 |
| `length`          | width of the segment in bp                                       |
| `is_exon`         | logical: is this a cassette exon?                                |
| `cod`, `cod.gene` | coding status of the segment (within CDS)                        |
| `ncell`           | number of single cells with coverage ≥10 for this segment        |
| `nna`             | how many pseudobulks had `NA` for psi (insufficient reads)       |
| `sd`              | standard deviation of ψ across pseudobulks                       |


```{r rowRanges, cache = TRUE}
SummarizedExperiment::rowRanges(pbasf)[1:3]
# View(pbasf@rowRanges)
```

<br><br>

### information about pseudobulks (per sample/celltype)
| Column            | Meaning                                                          |
| ----------------- | ---------------------------------------------------------------- |
| `sample_id`       | original sample ID                                               |
| `celltype`        | the celltype (cluster)                                           |
| `ncells`          | cells aggregated into the pseudobulk                             |
| `strand`          | strand orientation of the coverage                               |

```{r colData, cache = TRUE}
pbasf@colData
```

<br><br>

### assays
| Assay name | Contents                                                   |
| ---------- | ---------------------------------------------------------- |
| `i`        | inclusion junction counts                                  |
| `e`        | exclusion junction counts                                  |
| `psi`      | percent-spliced-in $\frac{i}{(i+e)}$;  NA if $i+e<10$      |

```{r assays, cache = TRUE}
pbasf@assays
SummarizedExperiment::assay(pbasf, "i")[1:5, 1:4]
SummarizedExperiment::assay(pbasf, "e")[1:5, 1:4]
SummarizedExperiment::assay(pbasf, "psi")[1:5, 1:4]
```

<br><br>

### metadata: postprocessing results
- `all_celltype_test`
  - `group`/`group_fdr` — test statistics & FDR
  - `low_state`, `high_state` — celltype with lowest and highest
  - `dpsi` — difference in psi between groups
- `markers`
  - `pv` — p-value per segment and celltype
  - `fdr` — FDR per segment and celltype
  - `dpsi` — difference in psi per segment and celltype
- `go` — GO enrichment results
- `ipro` — domain enrichment results

```{r metadata, cache = TRUE}
names(pbasf@metadata)
head(S4Vectors::metadata(pbasf)$all_celltype_test)
str(S4Vectors::metadata(pbasf)$markers)
str(S4Vectors::metadata(pbasf)$go)
str(S4Vectors::metadata(pbasf)$ipro)
```

<br><br>


# Find differentially spliced segments

## Results comparing all celltypes simultaneously $\text{fdr} < 0.05$
<!-- Need to understand statistics -->
```{r all_celltype_test, eval=FALSE, cache = TRUE}
subset(
  pbasf@metadata$all_celltype_test,
  !is.na(dpsi) & group_fdr < 0.05
)
```

## Results of comparing each celltype against the rest are stored as three (pv, fdr, dpsi) $\text{segment} * \text{celltype}$ matrices
```{r markers, cache = TRUE}
pbasf@metadata$markers$fdr[1:2, ]
pbasf@metadata$markers$dpsi[1:2, ]
```


## Marker segment selection
markers: $\text{fdr} < 0.05$, $|\text{dpsi}| > 0.5$, at most 2 segments per celltype
```{r select_markers, eval=FALSE, cache = TRUE}
# select_markers only uses output of per-celltype tests
# let's find at most two marker segments per celltype, ensuring that each segment is reported only once
scsajr::select_markers(pbasf@metadata$markers, n = 2, dpsi_thr = 0.5, clean_duplicates = TRUE)
```

```{r selectAllMarkers, eval=FALSE, cache = TRUE}
# select_all_markers also takes results of "all celltypes together" test
# it gives priority to markers (they are marked by is_marker field)

# select_all_markers relies on select_markers and select_markers_from_all_celltype_test
scsajr::select_all_markers(pbasf@metadata$markers, pbasf@metadata$all_celltype_test, dpsi_thr = 0.5, n = Inf)
```


## Load gene expression data
This dataset is targeted, meaning that some neuronal genes were enriched, but other genes are still present. 
The AS analysis was run on all genes.

```{r Load gene expression data, cache = TRUE}
ge <- Seurat::Read10X_h5("./input/raw_feature_bc_matrix.h5", use.names = FALSE) # Load 10x h5 file
ge <- Seurat::CreateSeuratObject(ge) # Wrapping into Seurat object
```

```{r Subset gene expression data, cache = TRUE}
ge <- ge[, barcodes$barcode] # Keep only cells annotated in barcodes.tsv
ge$barcode <- colnames(ge) # Add barcode as a column
colnames(ge) <- paste0("sample|", colnames(ge)) # Rename column names to include sample_id in barcodes
ge$celltype <- barcodes[colnames(ge), "celltype"] # Add pre-computed celltype labels from barcodes.tsv
```

```{r Normalisation, warning=FALSE, cache = TRUE}
# Just use standard Seurat pipeline to re-cluster cells
ge <- Seurat::NormalizeData(ge) # Log-normalise UMI counts
ge <- Seurat::FindVariableFeatures(ge) # Find highly variable features
# VariableFeatures(ge)
ge <- Seurat::ScaleData(ge)
```

```{r PCA and UMAP, message=FALSE, cache = TRUE}
ge <- Seurat::RunPCA(ge)
ge <- Seurat::FindNeighbors(ge)
ge <- Seurat::FindClusters(ge)
ge <- Seurat::RunUMAP(ge, dims = 1:30)
Seurat::DimPlot(ge, reduction = "umap", group.by = c("celltype", "seurat_clusters"))

# Here we have just one sample, lets add sample_id to make code compatible with multiple sample design
ge$sample_id <- "sample"
```

# Convert Seurat to SummarizedExperiment & pseudobulk expression
```{r Convert Seurat to SummarizedExperiment, cache = TRUE}
genes <- gene_descr[rownames(ge), ]
rowRanges <- GenomicRanges::GRanges(genes$chr,
  IRanges::IRanges(start = genes$start, end = genes$end),
  strand = ifelse(genes$strand == 1, "+", ifelse(genes$strand == -1, "-", "*")),
  feature_id = rownames(genes)
)
genes$chr <- genes$start <- genes$end <- genes$strand <- NULL

S4Vectors::elementMetadata(rowRanges)[names(genes)] <- genes

sce <- SummarizedExperiment(
  assays = list(counts = ge[["RNA"]]$counts),
  rowRanges = rowRanges,
  colData = ge@meta.data
)

pbge <- scsajr::pseudobulk(sce, c("sample_id", "celltype")) # calculate pseudobulk by celltype
SummarizedExperiment::assay(pbge, "cpm") <- scsajr::calc_cpm(pbge)
```

## Visualise markers
```{r Visualise markers, fig.width=8, fig.height=8, cache = TRUE}
markers <- scsajr::select_all_markers(
  pbasf@metadata$markers, pbasf@metadata$all_celltype_test,
  dpsi_thr = 0.5, n = 4
)

par(oma = c(0, 14, 0, 2), mar = c(5, 9, 3, 4))
# par(mar = c(0, 0.5, 0, 0), oma = c(6, 34, 3, 1), xpd = NA)

scsajr::marker_heatmap(
  pbasf, pbge, "celltype",
  psi_scale = FALSE,
  cpm_scale = TRUE,
  markers = markers,
  gene_names_col = "name"
)
```

## Coverage plots
From pre-saved files, we can replot coverage for one of examples generated by the pipeline

```{r Coverage plots, cache = TRUE}
sid <- "ENSG00000092841.s17"
covs <- readRDS(paste0("./nf-scsajr_output/rds/examples_coverage/", sid, ".rds"))
scsajr::plot_segment_coverage(
  sid,
  chr = covs[[1]]$chr,
  start = covs[[1]]$start,
  stop = covs[[1]]$end,
  covs = covs,
  data_as = pbasf,
  data_ge = pbge,
  groupby = "celltype",
  barcodes = barcodes,
  samples = samples,
  gene_descr = gene_descr,
  plot_junc_only_within = FALSE,
  ylim_by_junc = TRUE,
  gtf = gtf,
  oma = c(4, 4, 4, 1)
)
```


We can zoom in on the alternative exon and focus on the most divergent cell types.
```{r Zoom in on alternative exon, cache = TRUE}
scsajr::plot_segment_coverage(
  sid,
  chr = covs[[1]]$chr,
  start = 56160210,
  stop = 56161500,
  covs = covs,
  data_as = pbasf,
  data_ge = pbge,
  celltypes = c("6", "9", "5"),
  groupby = "celltype",
  barcodes = barcodes,
  samples = samples,
  gene_descr = gene_descr,
  plot_junc_only_within = FALSE,
  ylim_by_junc = TRUE,
  gtf = gtf,
  oma = c(8, 4, 4, 1)
)
```


## Differential splicing
Let's compare just two celltypes, we will take cluster 5 and 6 as they demonstrated divergent splicing in MYL6
```{r Differential splicing, message = FALSE, warning= FALSE, cache = TRUE}
cl5to6 <- scsajr::test_pair_as(pbasf, "celltype", c("5", "6"))
cl5to6[!is.na(cl5to6$dpsi) & cl5to6$fdr < 0.05 & abs(cl5to6$dpsi) > 0.9, ]
```

```{r Inspect segment and plot, cache = TRUE}
sid <- "ENSG00000217930.s9"
pbasf@metadata$all_celltype_test[sid, ]
pbasf@metadata$markers$fdr[sid, ]

# _coverage plots from BAM file ####
# when segment of interest was not plotted by pipeline
# or if celltype annotation has changed, one may need to plot coverage from BAM files
covs <- NULL # plot_segment_coverage returns coverage object that can be used next time
covs <- scsajr::plot_segment_coverage(
  sid,
  start = 4343250,
  stop = 4351500,
  covs = covs,
  data_as = pbasf,
  data_ge = pbge,
  groupby = "celltype",
  barcodes = barcodes,
  samples = samples,
  gene_descr = gene_descr,
  celltypes = c("5", "6"),
  plot_junc_only_within = FALSE,
  ylim_by_junc = TRUE,
  gtf = gtf,
  oma = c(4, 4, 4, 1)
)
# Exon seems to be included in cluster 5 and excluded in cluster 6. However, coverage is quite low
```

```{r Save coverage for future use, cache = TRUE}
# We can save coverage for future use.
# It is not very important in case of single sample, but with tens of samples and celltypes extracting coverage from BAM files can take a while.
saveRDS(covs, paste0("./nf-scsajr_output/rds/examples_coverage/", sid, ".rds"))
```

## domain enrichment analyses
```{r Domain enrichment analyses, cache = TRUE}
# Load segment to domain annotation and domain description
domain2seg <- readRDS(url("https://github.com/cellgeni/nf-scsajr/raw/refs/heads/main/ref/human_2020A/functional_annotation/domain2seg.df.rds"))
domain_descr <- readRDS(url("https://github.com/cellgeni/nf-scsajr/raw/refs/heads/main/ref/human_2020A/functional_annotation/all_domain_descr.rds"))

# Load all segments
pbas_all <- readRDS("./nf-scsajr_output/rds/pbas.rds")

# Take all domains that have description
domain2seg <- domain2seg$all
domain2seg <- domain2seg[domain2seg$domain %in% domain_descr$ENTRY_AC, ]

domain_sites2use <- "ad" # focus on cassette exons only
# Select background set
# All genes that have at least one segment with reasonable coverage
gene_uni <- scsajr::filter_segments_and_samples(pbas_all, seg_min_sd = 0, celltype_min_samples = 1, sample_min_ncells = 30)
gene_uni <- unique(rowData(gene_uni)$gene_id)
seg_uni <- rownames(pbas_all)[segs$is_exon & segs$sites %in% domain_sites2use & segs$gene_id %in% gene_uni]

cl5to6sgn <- cl5to6[!is.na(cl5to6$dpsi) & cl5to6$fdr < 0.05 & abs(cl5to6$dpsi) > 0.1, ]
sids <- list(
  cl6 = rownames(cl5to6sgn)[cl5to6sgn$dpsi > 0.1],
  cl5 = rownames(cl5to6sgn)[cl5to6sgn$dpsi < -0.1]
)

ipro <- clusterProfiler::compareCluster(sids,
  fun = "enricher",
  universe = seg_uni,
  pAdjustMethod = "BH",
  TERM2GENE = domain2seg,
  TERM2NAME = domain_descr[, c("ENTRY_AC", "ENTRY_NAME")]
)

clusterProfiler::dotplot(ipro)
```

## re-pseudobulk
```{r Re-pseudobulk, cache = TRUE}
# Using new cell clusters
# redo pbasf, pbge, barcodes
raw <- readRDS("./nf-scsajr_output/rds/sample.rds")
barcodes_new <- ge@meta.data[c("sample_id", "barcode", "seurat_clusters")]

raw$i <- raw$i[, rownames(barcodes_new)]
raw$e <- raw$e[, rownames(barcodes_new)]
raw$seg <- segs
pbas_new <- scsajr::make_summarized_experiment(raw, barcodes_new)
pbas_new <- scsajr::pseudobulk(pbas_new, "seurat_clusters")
pbasf_new <- scsajr::filter_segments_and_samples(
  pbas_new,
  groupby = "seurat_clusters",
  seg_min_sd = 0.1,
  celltype_min_samples = 1,
  sample_min_ncells = 30
)
# str(pbasf_new)
```